---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, we load the data into a data frame, called "activity". Using str, we can see what the raw data looks like.

```{r load}
activity <- read.csv(unz("activity.zip", "activity.csv"))
str(activity)
```

Next, we will add a column called "timestamp" that converts the date and interval variables into a single posixct object for each observation. First, we'll need to transform the interval variable to strings of matching length. 0 will become "0000", 5 "0005", up to 2355 to "2355".
```{r preprocess}

##transform interval to format "0000", "0005", "0010", etc.
##format = "d" indicates interval is currently an integer, 
##and flag appends 0s as necessary to the front of the string.
activity$interval <- 
      formatC(activity$interval, width = 4, format = "d", flag = "0")

##paste date and interval, and convert to date object using strptime
activity$timestamp <- with(activity, paste(date, interval))
activity$timestamp <- strptime(activity$timestamp, format = "%Y-%m-%d %H%M")
```

## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
