---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, we load the data into a data frame, called "activity". Using str, we can see what the raw data looks like.

```{r load}
activity <- read.csv(unz("activity.zip", "activity.csv"))
str(activity)
```

Next, we will add a column called "timestamp" that converts the date and interval variables into a single posixct object for each observation. First, we'll need to transform the interval variable to strings of matching length. 0 will become "0000", 5 "0005", up to 2355 to "2355".
```{r preprocess}

##transform interval to format "0000", "0005", "0010", etc.
##format = "d" indicates interval is currently an integer, 
##and flag appends 0s as necessary to the front of the string.
activity$interval <- 
      formatC(activity$interval, width = 4, format = "d", flag = "0")

##paste date and interval, and convert to date object using strptime
activity$timestamp <- with(activity, paste(date, interval))
activity$timestamp <- strptime(activity$timestamp, format = "%Y-%m-%d %H%M")
```

## What is mean total number of steps taken per day?

For this we will make a histogram of the daily totals for steps using the base plotting system, along with lines showing the mean and median number of steps. These calculations will ignore days for which there are no observed step data.

```{r histogram}
##plot histogram with titles, mean and median lines, and legend
hist(with(activity, tapply(steps, date, sum)), breaks = 10,
     main = "Histogram of Daily Total Steps", xlab = "Steps")
abline(v = mean(with(activity, tapply(steps, date, sum)), na.rm = TRUE), 
       col = "blue")
abline(v = median(with(activity, tapply(steps, date, sum)), na.rm = TRUE), 
       col = "red")
legend("topright", lty=c(1,1), col = c("blue", "red"), 
       legend = c("Mean", "Median"))

##calculate mean and median for output below
stepMean <- mean(with(activity, tapply(steps, date, sum)), na.rm = TRUE)
stepMedian <-median(with(activity, tapply(steps, date, sum)), na.rm = TRUE)

```

As shown on the histogram, the mean and median daily step totals are very close. The mean number of steps is `r stepMean`. The median number of steps is `r stepMedian`.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
